---
title: "Diffsig: Associating Risk Factors With Mutational Signatures"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
author: "Ji-Eun Park, Michael Love"
output: 
  rmarkdown::html_document:
    self_contained: false
    highlight: pygments
abstract: |
  Diffsig provides estimation of the associations between mutational signatures
  and risk factors to help better understand the cancer development process. 
  Diffsig generates the associations using a hierarchical Bayesian
  Dirichlet-Multinomial model with a set of mutational signatures, mutation 
  counts, and risk factors. Diffsig can test posterior estimates of the 
  associations with any type of discrete or continuous risk factors of interest
  while considering the possible variance derived by datasets sequenced with 
  various read depth.
  Diffsig package version: `r packageVersion("diffsig")`
bibliography: library.bib
vignette: |
  %\VignetteIndexEntry{Diffsig - estimation of association of risk factors and 
  mutational signatures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

*Diffsig* provides Bayesian analysis of the association between risk
factors and mutational signatures, making use of the Stan and RStan
software packages for inference [@stan; @rstan].

The authors of the associated [Diffsig manuscript]() [@diffsig] are:

> Ji-Eun Park\*, Markia Smith, Sarah Van Alsten, Andrea Walens, Di Wu,
> Katherine A. Hoadley, Melissa A. Troester, Michael I. Love

```{r}
library(diffsig)
```


# Data simulation

The purpose of this vignette is to demonstrate the steps of *Diffsig*.
For that, in this vignette, we work on simulated mutational counts and
risk factors instead of real datasets. The simulation process does not
include steps of generating mutational count matrix from .vcf or .maf
files. Details for the count matrix generating process could be found in
the manuscript [public data preprocess]().

A number of parameters need to be set to generate a simulated dataset: +
`N`: number of samples (integer) + `K`: number of signatures (integer) +
`truetau`: true value of tau (numeric) + `C_k`: signature matrix (96xK
matrix)

In our analysis, we used the COSMIC V3.2 signatures loaded from
[COSMIC](https://cancer.sanger.ac.uk/signatures/sbs/) where we selected
5 signatures that are known to be associated with breast cancer to avoid
selecting an arbitrary set of signatures. If you have your own signature
matrix, you can load your signature matrix object named as `C_k`.

```{r}
N = 50
K = 5
truetau = 100
C = read.table("https://cancer.sanger.ac.uk/signatures/documents/453/COSMIC_v3.2_SBS_GRCh38.txt", header = T, row.names = 1)
C_k =  C[,c("SBS1","SBS2","SBS3","SBS5","SBS13")]
```

Now, there are some additional parameters that should be set depending
on the types of risk factors to be included:

-   continuous: number of continuous variables (default `NULL`)
-   binary: number of binary variables (default `NULL`)
-   categorical: number of categorical variables (default `NULL`)
-   cat_levels: vector of number of levels for each categorical
    variables (default `NULL`)

So, for example, if you would like to include 2 continuous variable, you
would do:

```{r}
set.seed(123)
simdat = simdat_gen(N=N, K=K, truetau=truetau, C=C_k,
           continuous=2, binary=NA, categorical=NA, cat_levels=NA, seed=123)
```

Or, if you want to include 1 binary and 2 categorical variable with 3
and 4 levels:

```{r, eval=F}
simdat = simdat_gen(N=N, K=K, truetau=truetau, C=C_k, 
           continuous=NA, binary=1, categorical=2, cat_levels=c(3,4))
```

Note that `continuous`, `binary`, `categorical`, `cat_levels` are set to
`NULL` as default. Therefore, you do not need to specify anything for
variables you wish not to include.

```{r, eval=F}
# Same as the above examples
simdat_gen(N=N, K=K, truetau=truetau, C=C_k, continuous=2)
simdat_gen(N=N, K=K, truetau=truetau, C=C_k, binary=1, categorical=2, cat_levels=c(3,4))
```

The `simdat` object includes the three components: true association
`beta`, mutational counts `Y`, and risk factor matrix `X`.

```{r}
head(simdat$truebeta)
```

```{r}
head(simdat$Y)
```

```{r}
head(simdat$X)
```

# Association estimation

Once we obtain the mutation counts `Y`, risk factors `X`, and mutational
signatures matrix `C`/`C_K`, we can estimate the association between
risk factors and mutational signatures using the function `diffsig_fit`.

```{r}
fit <- diffsig_fit(X=simdat$X, Y=simdat$Y, C=C_k, beta_sd=0.5,
                   chains=4, cores=4, seed=123)
```

This runs the hierarchical Bayesian model that yields the posterior
estimation of the associations between the risk factors and mutational
signature `beta`. In our analysis, we use 4 chains and 4 cores with 2000
iterations.

Depending on your interest, sample-signature association from Dirichlet
sampling `phi` and the precision parameter `tau` could be extracted from
the model by including `pars=c("beta","tau","phi")` in `diffsig_fit()`.

# Convergence examination

In order to check if the model converged, we recommend to check the
R-hat convergence measure.[@stan; @rstan] This diagnostic measure shows
how well the posterior estimates are converging and mixing in split
chains. `rstan` recommends to use samples with R-hat values less than
1.05 and it seems that our simulation has a well-converged result with
all R-hats below 1.01. `traceplots` from `rstan` is another common way
to check convergence. The `traceplot()` function shows how well the
chains are mixing across chains, where a poorly mixed result will show
divergence between chains. Traceplots from our simulated data shows a
nicely mixed and well-converged result across all 4 chains.

```{r}
mu_tau_summary <- summary(fit, pars = c("beta"), probs = c(0.1, 0.9))$summary
data.frame(Rhat =mu_tau_summary[,"Rhat"])
rstan::traceplot(fit, pars="beta")
```

Additionally, *Diffsig* provides a function `diffsig_eval` to generate
evaluation for simulated data results. The function outputs a list of
three objects: summary statistics matrix `estimate` with the desired
credible interval width and two measurements `perc_coverage` and
`mean_rmse`. + `perc_coverage`: percentage of true values belonging
within credible interval + `mean_rmse`: average RMSE of each estimated
`beta` values\
The width of the credible interval can be specified with
`credibleInterval` where in default is set to
80%(`ci_level=80`).

```{r}
(eval_res = diffsig_eval(betatrue = simdat$truebeta, fit = fit, ci_level = 80))
```

It is ideal to have results with coverage greater than 80% or mean RMSE
less than 0.1. In our example, we can see that the coverage percentage
is `r round(eval_res$perc_coverage,2)*100`% and mean RMSE around
`r round(eval_res$mean_rmse,2)` which means that we acquired a good
estimation. Coverage and RMSE can be improved by increasing the sample
size which increases the power of the model.

# Plotting associations

Once we observe good convergence, we can then interpret the estimations
we obtained by plotting the posterior estimates. `diffsig_plot()` is a
modified version of the `stan_plot()` function from the `rstan` package
which allows to group the segments by groups (or risk factors) for
easier visualization. Since each of the estimations are for the `K`
signatures and `M` risk factors, the rows are labelled with the
signatures with `signature_labels` and are divided and colored for each
risk factor with `riskfactor_labels`.

```{r}
numpars = prod(dim(simdat$truebeta)) #15
diffsig_plot(fit, 
             pars=1:numpars, 
             ci_level = 80,
             signature_labels = c("SBS1","SBS2","SBS3","SBS5","SBS13"), 
             riskfactor_labels = c("intercept","cont1","cont2"),
             ncol=1, outlist = F)
```

Each row shows the estimated association between each signature and risk
factor. The circle points represent the point estimations, black lines
represent the credible intervals (80% CI in this case), and the colored
segments represent the 25% and 75% quantile, respectively.

In addition,

-   `ncol`: by default is set to 1, which indicates the plots for each
    risk factor will be printed vertically in 1 column. If `ncol` is set
    to the number of risk factors (including intercept), plots will be
    printed horizontally. `ncol` must be an integer between 1 to the
    number of risk factors.
-   `outlist`: by default is set to `FALSE`, which indicates the final
    output is a single plot with all plots aligned based on the
    specified `ncol`. If set to `TRUE`, the output will be a list of
    individual plots for each risk factor.

```{r}
diffsig_plot(fit, 
             pars=1:numpars, 
             ci_level = 80,
             signature_labels = c("SBS1","SBS2","SBS3","SBS5","SBS13"), 
             riskfactor_labels = c("intercept","cont1","cont2"),
             ncol=3, outlist = F)
```

```{r}
diffsig_plot(fit, 
             pars=1:numpars, 
             ci_level = 80,
             signature_labels = c("SBS1","SBS2","SBS3","SBS5","SBS13"), 
             riskfactor_labels = c("intercept","cont1","cont2"),
             ncol=1, outlist = T)
```

The colors for the segments and estimation points can also be tailored
based on preference by specifying `colors` and `est_color` parameters.

```{r}
diffsig_plot(fit, 
             pars=1:numpars, 
             ci_level = 80,
             signature_labels = c("SBS1","SBS2","SBS3","SBS5","SBS13"), 
             riskfactor_labels = c("intercept","cont1","cont2"),
             colors=c("#D81B60","#1E88E5","#FFC107"),
             est_color="#F5F3EF",
             ncol=1,
             outlist=FALSE)
```

# Interpreting the results

Note that estimates must be interpreted within each risk factor and not
be compared between estimates for different risk factors. For example,
let's call the association between SBS1 and SBS2 with risk factor 1 as
$\beta_{11}$ and $\beta_{12}$, respectively, and association between
SBS1 and SBS2 with risk factor 2 as $\beta_{21}$ and $\beta_{22}$
,respectively. From the result we can see that
$\beta_{11}=$`r round(eval_res[[1]][6,2],2)`,
$\beta_{12}=$`r round(eval_res[[1]][7,2],2)`,
$\beta_{21}=$`r round(eval_res[[1]][11,2],2)`, and
$\beta_{22}=$`r round(eval_res[[1]][12,2],2)`. What we can compare is
only between ($\beta_{11}$ and $\beta_{12}$) or between ($\beta_{21}$
and $\beta_{22}$), e.g. $\beta_{11}<\beta_{12}$ so risk factor 1 has
higher association with SBS2 than 1. What we cannot compare is across
($\beta_{11}$,$\beta_{12}$) and ($\beta_{21}$,$\beta_{22}$), e.g. we
cannot say that risk factor 1 is more associated with SBS2 because
$\beta_{12}>\beta_{22}$. Reasons for such limitation can be found from
the Discussion of our paper (link)

# Session info

```{r}
sessionInfo()
```

# References
