---
title: "Diffsig - "
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
author: "Ji-Eun Park, Michael Love"
output: 
  rmarkdown::html_document:
    self_contained: false
    highlight: pygments
abstract: |
  Diffsig provides estimation of the associations between mutational signatures
  and risk factors to help better understand the cancer development process. 
  Diffsig generates the associations using a hierarchical Bayesian
  Dirichlet-Multinomial model with a set of mutational signatures, mutation 
  counts, and risk factors. Diffsig can test posterior estimates of the 
  associations with any type of discrete or continuous risk factors of interest
  while considering the possible variance derived by datasets sequenced with 
  various read depth.
  Diffsig package version: `r packageVersion("diffsig")`
bibliography: library.bib
vignette: |
  %\VignetteIndexEntry{Diffsig - estimation of association of risk factors and 
  mutational signatures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

*Diffsig* provides Bayesian analysis of the association between risk
factors and mutational signatures, making use of the Stan and RStan
software packages for inference [@stan; @rstan].

The authors of the associated [Diffsig manuscript]() [@diffsig] are:

> Ji-Eun Park\*, Markia Smith, Sarah Van Alsten, Andrea Walens, Di Wu,
> Katherine Hoadley, Melissa Troester, Michael I. Love

# Data simulation

The purpose of this vignette is to demonstrate the steps of *Diffsig*.
For that, in this vignette, we work on simulated mutational counts and
risk factors instead of real datasets. The simulation process does not
include steps of generating mutational count matrix from .vcf or .maf
files. Details for the count matrix generating process could be found in
the manuscript [public data preprocess]().

A number of parameters need to be set to generate a simulated dataset:
+ `N`: number of samples (integer) 
+ `K`: number of signatures (integer) 
+ `truetau`: true value of tau (numeric) 
+ `C_K`: signature matrix (96xK matrix)

In our analysis, we used the COSMIC V3.2 signatures loaded from
[COSMIC](https://cancer.sanger.ac.uk/signatures/sbs/) where we selected 5 signatures that are known to be associated with breast cancer to avoid selecting an arbitrary set of signatures. If you have your own signature matrix, you can load your signature matrix object named as `C_K`. 

```{r}
N = 50
K = 5
truetau = 100
C = read.table("https://cancer.sanger.ac.uk/signatures/documents/453/COSMIC_v3.2_SBS_GRCh38.txt", header = T, row.names = 1)
C_K =  C[,c("SBS2","SBS3","SBS5","SBS8","SBS13")]
```

Now, there are some additional parameters that should be set depending
on the types of risk factors to be included:

-   continuous: number of continuous variables (default `NULL`)
-   binary: number of binary variables (default `NULL`)
-   categorical: number of categorical variables (default `NULL`)
-   cat_levels: vector of number of levels for each categorical
    variables (default `NULL`)

So, for example, if you would like to include 2 continuous variable, you
would do:

```{r}
simdat = simdat_gen(N=N, K=K, truetau=truetau, C=C_K,
           continuous=2, binary=NA, categorical=NA, cat_levels=NA, seed=123)
```

Or, if you want to include 1 binary and 2 categorical variable with 3
and 4 levels:

```{r, eval=F}
simdat = simdat_gen(N=N, K=K, truetau=truetau, C=C, 
           continuous=NA, binary=1, categorical=2, cat_levels=c(3,4))
```

Note that `continuous`, `binary`, `categorical`, `cat_levels` are set to
`NULL` as default. Therefore, you do not need to specify anything for
variables you wish not to include.

```{r, eval=F}
# Same as the above examples
simdat_gen(N=N, K=K, truetau=truetau, C=C, continuous=2)
simdat_gen(N=N, K=K, truetau=truetau, C=C, binary=1, categorical=2, cat_levels=c(3,4))
```

The `simdat` object includes the three components: true association `beta`,
mutational counts `Y`, and risk factor matrix `X`. 

```{r}
head(simdat$truebeta)
```

```{r}
head(simdat$Y)
```

```{r}
head(simdat$X)
```

# Association estimation 

Once we obtain the mutation counts `Y`, risk factors `X`, and mutational signatures matrix `C`/`C_K`, we can estimate the association between risk factors and mutational signatures using the function `diffsig_fit`.

```{r}
fit <- diffsig_fit(X=simdat$X, Y=simdat$Y, C=C_K, beta_sd=0.5,
                   chains=4, cores=4, seed=123)
```

This runs the hierarchical Bayesian model that yields the posterior estimation of the associations between the risk factors and mutational signature `beta`. In our analysis, we use 4 chains and 4 cores with 2000 iterations.

Depending on your interest, sample-signature association from Dirichlet sampling `phi` and the precision parameter `tau` could be extracted from the model by including `pars=c("beta","tau","phi")` in `diffsig_fit()`. 

# Convergence examination

In order to check if the model converged, we recommend to check the R-hat convergence measure.[@stan; @rstan] This diagnostic measure shows how well the posterior estimates are converging and mixing in split chains. `rstan` recommends to use samples with R-hat values less than 1.05 and it seems that our simulation has a well-converged result with all R-hats below 1.01. `traceplots` from `rstan` is another common way to check convergence. The `traceplot()` function shows how well the chains are mixing across chains, where a poorly mixed result will show divergence between chains. Traceplots from our simulated data shows a nicely mixed and well-converged result across all 4 chains.

```{r}
mu_tau_summary <- summary(fit, pars = c("beta"), probs = c(0.1, 0.9))$summary
data.frame(Rhat =mu_tau_summary[,"Rhat"])
rstan::traceplot(fit, pars="beta")
```

Additionally, *Diffsig* provides a function `diffsig_eval` to generate evaluation for simulated data results. The function outputs a list of three objects: summary statistics matrix `estimate` with the desired credible interval width and two measurements `perc_coverage` and `mean_rmse`.
+ `perc_coverage`: percentage of true values belonging within credible interval
+ `mean_rmse`: average RMSE of each estimated `beta` values  
The width of the credible interval can be specified with `credibleInterval` where in default is set to 80%(`credibleInterval=80`). 

```{r}
(eval_res = diffsig_eval(truebeta = truebeta, fit = fit, ci_level = 80))
```

It is ideal to have results with coverage greater than 80% or mean RMSE less than 0.1. In our example, we can see that the coverage percentage is `r round(eval_res$perc_coverage,2)*100`% and mean RMSE around `r round(eval_res$mean_rmse,2)` which means that we acquired a good estimation. Coverage and RMSE can be improved by increasing the sample size which increases the power of the model.

# Plotting associations

Once we observe good convergence, we can then interpret the estimations we obtained by plotting the posterior estimates. `diffsig_plot()` is a modified version of the `stan_plot()` function from the `rstan` package which allows to group the segments by groups (or risk factors) for easier visualization. Since each of the estimations are for the `K` signatures and `M` risk factors, the rows are repeatedly labelled with the signatures with `rowlabels` and are colored based on risk factors with `rowgroup`.

```{r}
diffsig_plot(fit, pars=1:15, ci_level = 80,
             rowlabels = rep(c("SBS2","SBS3","SBS5","SBS8","SBS13"),3), 
             rowgroup = rep(c("intercept","cont1","cont2"),each=5))
```
Each row shows the estimated association between each signature and risk factor. The circle points represent the point estimations, black lines represent the credible intervals (80% CI in this case), and the colored segments represent the 25% and 75% quantile, respectively.

**FIX THIS PART** Note that.. relativity.. **FIX THIS PART**

The colors for the segments and estimation points can also be tailored based on preference by specifying `colors` and `est_color` parameters.

```{r}
diffsig_plot(fit, pars=1:15, ci_level = 80,
             rowlabels = rep(c("SBS2","SBS3","SBS5","SBS8","SBS13"),3), 
             rowgroup = rep(c("intercept","cont1","cont2"),each=5),
             colors=c("#D81B60","#1E88E5","#FFC107"),
             est_color="#F5F3EF") +
  geom_point(aes(y=15:1,x=simdat$truebeta),shape=3)
```

# Session info

```{r}
sessionInfo()
```

# References
